The Design and Implementation of the Synchronous Language \CEU

### 2016-03-29 paper.tex
...

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction to events.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulates for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO: Such analsysis is orthogonal

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
Howerver, \CEU supports internal events which programs can \code{emit} in 
addition to \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then, periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these specific corner 
cases to the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages, such as 
\emph{SOL}~\cite{wsn.sol},
\emph{SC}~\cite{rp.sychc}, and
\emph{PRET-C}~\cite{rp.pretc},
enforce intra-reaction determinism with an arbitrary execution order for 
statements in multiple lines of execution.
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.

@@

Even so, we consider that enforcing an arbitrary execution order can be 
misleading in some cases.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
which define a shared variable (ln. 2), and assign to it parallel trails (ln.  
5, 8).
%
In the example \code{(a)}, the two assignments to \code{x} execute from 
reactions to different events \code{A} and \code{B} which, by definition, 
cannot occur simultaneously (Section~\ref{sec.ceu.exts}).
Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
for \code{B->A}, \code{x} becomes \code{(1*2)+1->3}.
%
In the example \code{(b)}, the two assignments to \code{y} execute from a 
reaction to the same event \code{A}, and are simultaneous from an external 
point of view.
Since \CEU employs lexical order for intra-reaction statements, the execution 
is still deterministic, and \code{y} always becomes \code{(1+1)*2->4}.
%
However, any (apparently innocuous) change in the order of trails can change 
the semantics of a program, which we consider unsafe.

To mitigate this threat, \CEU performs a temporal analysis at compile time and 
detects concurrent accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} in lines 2 and 5 in \code{(a)}
      \textbf{cannot} be concurrent because they are not in parallel trails.
\item The assignments to \code{x} in lines 5 and 8 in \code{(a)}
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} in lines 5 and 8 in \code{(b)}
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The algorithm for the analysis, which is depicted in 
Section~\ref{sec.impl.analsyis}, inspects all possible \code{await} statements 
that precede a variable access and hold the associated awaking events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

Note that this analysis is only possible due to the uniqueness of input events 
within reactions.
Otherwise, any two trail segments in parallel could be concurrent, even if they 
react to different input events.

@@

In \CEU, any identifier prefixed with an underscore is repassed \emph{as is} to 
the \C compiler that generates the final binary.
Therefore, access to \C is seamless and, more importantly, easily trackable.
%
Similarly to Esterel, which supports the \code{call} primitive for external 
code, calls are assumed to be instantaneous~\cite{esterel.primer}.
%
Evidently, programs should only resort to \C for asynchronous functionality, 
such as non-blocking I/O, or simple \code{struct} accessors, but never for 
control purposes.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ TODO
\label{lst.c}
}
\end{figure}

In addition, \CEU also employs the temporal analysis for calls and accesses to 
external symbols in \C.
%
As an example, the program in Figure~\ref{lst.c}.a defines four external 
symbols inside a \code{native} block for standard declarations in \C (ln.  
1--6).
%
The \code{par/and} (ln. 7--11) creates two trails that react concurrently 
during the boot reaction (ln. 8,10):
the first trail calls the symbol \code{\_f},
while the second calls \code{\_g} and \code{\_id}, and also accesses 
\code{\_NUM}.
%
Since \CEU does not inspect any code defined in \C, it has no idea about the 
meaning of the symbols and complains about suspicious concurrent accesses 
between \code{\_f} against all symbols in the second trail.

@@

Esterel's \code{abort} and \CEU's \code{par/or} statements illustrate how 
synchronous languages can abort awaiting lines of execution without tweaking 
them with synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
%
Still, aborting lines of execution that deal with external resources is 
challenging because they may end up in an inconsistent state.
%
For this reason, Esterel and \CEU provide a \code{finalize} construct with a 
clause that executes even if the enclosing block aborts and does not terminate 
normally.

Figure~\ref{lst.fin} uses \code{lock} and \code{unlock} calls to represent an 
external resource.
The normal behavior is to \code{lock} the resource, perform some operations in 
subsequent reactions to input \code{A}, and then \code{unlock} the resource.
However, if the aborting input \code{B} occurs after the \code{lock} but before 
the reactions to \code{A}, we still want to call \code{unlock} to safely 
release the resource.
%
In Esterel and \CEU, the finalize clauses (ln. a:10 and b:5) are automatically 
called if the enclosing blocks (ln. a:3--1 and b:3--10) are externally aborted 
(ln. a:12 and b:12).

@@

\subsection{ Discussion }
\label{sec.ceu.dicsussion}

>>>>

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction of a given event.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulates for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO: Such analsysis is orthogonal

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with intra-reaction communication, \CEU supports internal events which 
programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these specific corner 
cases to the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages, such as 
\emph{SOL}~\cite{wsn.sol},
\emph{SC}~\cite{rp.synchc}, and
\emph{PRET-C}~\cite{rp.pretc},
enforce intra-reaction determinism with an arbitrary execution order for 
statements in multiple lines of execution.
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.

@@

Even so, we consider that enforcing an arbitrary execution order can be 
misleading in some cases.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
which define a shared variable (ln. 2), and assign to it in parallel trails 
(ln.  5, 8).
%
In the example \code{(a)}, the two assignments to \code{x} execute from 
reactions to different events \code{A} and \code{B} which, by definition, 
cannot occur simultaneously (Section~\ref{sec.ceu.exts}).
Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
for \code{B->A}, \code{x} becomes \code{(1*2)+1->3}.
%
In the example \code{(b)}, the two assignments to \code{y} are simultaneous 
because they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution 
is still deterministic, and \code{y} always becomes \code{(1+1)*2->4}.
%
However, an (apparently innocuous) change in the order of trails changes the 
semantics of the program, which we consider unsafe.

To mitigate this threat, \CEU performs a temporal analysis at compile time and 
detects concurrent accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} in lines 2 and 5 in \code{(a)}
      \textbf{cannot} be concurrent because they are not in parallel trails.
\item The assignments to \code{x} in lines 5 and 8 in \code{(a)}
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} in lines 5 and 8 in \code{(b)}
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The algorithm for the analysis, which is depicted in 
Section~\ref{sec.impl.analysis}, inspects all possible \code{await} statements 
that precede a variable access and keep a list with all corresponding awaking 
events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

Note that this analysis is only possible due to the uniqueness of input events 
within reactions.
Otherwise, any two trail segments in parallel can be concurrent, even if they 
react to different input events.

@@

In \CEU, any identifier prefixed with an underscore is repassed \emph{as is} to 
the \C compiler that generates the final binary.
Therefore, access to \C is seamless and, more importantly, easily trackable.
%
Similarly to Esterel, which supports the \code{call} primitive for external 
code, calls are assumed to be instantaneous~\cite{esterel.primer}.
%
This way, programs should only resort to \C for asynchronous functionality, 
such as non-blocking I/O, or simple \code{struct} accessors, but never for 
control purposes (the compiler can restrict the available symbols in \C).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{(a) Definition and use of external symbols}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

.
\end{lstlisting}
\centering\small{(b) Annotations for the external symbols in (a).}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The unsafe program in (a) only compiles with the annotations in (b).
\label{lst.c}
}
\end{figure}

We extend the temporal analysis of \CEU for calls and accesses to external 
symbols in \C.
%
As an example, the program in Figure~\ref{lst.c}.a defines four external 
symbols inside a \code{native} block with standard declarations in \C (ln.  
1--6).
%
The \code{par/and} (ln. 7--11) creates two trails that react concurrently 
during the boot reaction (ln. 8,10):
the first trail calls the symbol \code{\_f},
while the second calls \code{\_g} and \code{\_id}, and also accesses 
\code{\_NUM}.
%
Since \CEU does not inspect any code defined in \C, it complains about 
suspicious concurrent accesses between \code{\_f} and all symbols in the second 
trail.

@@

With the annotations in \code{(b)}, the program in \code{(a)} compiles 
successfully.

Esterel's \code{abort} and \CEU's \code{par/or} statements distinguish how 
synchronous languages can abort awaiting lines of execution without tweaking 
them with synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
%
Still, aborting lines of execution that deal with external resources is 
challenging because they may end up in an inconsistent state.
%
For this reason, Esterel and \CEU provide a \code{finalize} construct to 
unconditionally execute a series of statements even if the enclosing block is 
aborted and does not terminate normally.

The example in Figure~\ref{lst.fin} in Esterel (\code{(a)}) and \CEU 
(\code{(b)}) uses the \code{lock} and \code{unlock} calls which represent 
accesses to an external resource.
The normal behavior is to \code{lock} the resource (ln. \code{(a)}:4 and 
\code{(b)}:3), perform some operations in subsequent reactions to input 
\code{A} (ln. \code{(a)}:5--8 and \code{(b)}:7--10), and then \code{unlock} the 
resource (ln. \code{(a)}:10 and \code{(b)}:5).
However, if the aborting input \code{B} (ln. \code{(a)}:12 and \code{(b)}:12) 
occurs after the \code{lock} but before the reactions to \code{A}, we still 
want to call \code{unlock} to safely release the resource.
%
In Esterel and \CEU, the finalize clauses (ln. \code{(a)}:10 and \code{(b)}:5) 
are automatically called if the enclosing blocks (ln. \code{(a)}:3--1 and 
\code{(b)}:3--10) are externally aborted (ln. \code{(a)}:12 and \code{(b)}:12).

@@

\subsection{ Discussion }
\label{sec.ceu.discussion}
